Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool

### 2024-05-09 paper/PaperCEUS/PaperCEUS_prerev.tex

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

  \title{Reproducible methods for modeling combined public transport and
cycling trips and associated benefits: evidence from the biclaR tool}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit as a
  substitute for car trips in the Lisbon metropolitan area and assesses
  its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of
  a metropolitan cycling network was developed (\emph{biclaR}). The
  social and environmental impacts were assessed using Health World
  Organization tools. The impacts of shifting car trips to public
  transport were also estimated and monetized. The results show that
  10\% of all trips could be made by cycling in combination with public
  transport. Shifting to cycling for the shorter first and last mile
  stages can reduce annual CO\textsubscript{2}eq emissions from 3,000 to
  7,500 tons/day, while for the public transport leg, the transfer from
  car avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
  year. The estimated socio-environmental benefits are of €125 million
  to €325 million over 10 years. This evidence can support policymakers
  to prioritize interventions that reduce the reliance on private motor
  vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326, vanmil_insights_2021}. In The Netherlands,
which has the highest mode share of cycling in the world, cycling
accounts for more than a third of all trips to and from rail stations at
the `home' end of the journey, greatly increasing the ability of the
transport system \citep{RIETVELD200071}. This approach to reducing car
dependency and associated externalities requires interventions and
programs to make bicycling more appealing \citep{lapaix_role_2021}. The
resulting public investments can have significant social and
environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high.

The Portuguese Government has established a national cycling strategy,
which outlines targets for the percentage of trips to be made by bicycle
in urban areas. Specifically, these targets aim for 4\% of all trips in
2025 and 10\% by 2030 \citep{ENMAC}. Moreover, the strategy emphasizes
that this increase should be achieved by substituting car trips for
bicycle trips, as a way to reduce the environmental and social impacts
of car use.

To support the implementation of this strategy, the Lisbon's
Metropolitan Department of Transport commissioned
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network \citep{felix2023}.

@@

It has become progressively more common to establish strategic plans, at
national, regional or municipal level, to mitigate climate change. Among
these, the Sustainable Urban Mobility Plans (SUMP\footnote{See
  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
promoted by the European Commission, are becoming popular in Europe,
although authorities are designing documents of this sort all over the
world. The definition of targets associated with a timeframe for
reducing dependence on the individual motorized vehicle, or targets for
the use of active modes such as walking and cycling, are too often not
accompanied by estimates of their social, environmental and economic
impacts. It is important for authorities and practitioners to know how
to estimate those impacts, which tools are available to support them in
the process, and what results to expect.

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA. After presenting the
\hyperref[methods]{Methods} used in Chapter 2, we assess its
socio-environmental impacts using open data and open-source tools,
presenting the \hyperref[results-and-discussion]{Results and Discussion}
in Chapter 3, followed by Chapter 4 where we discuss the
\hyperref[generalizability-and-limitations]{Generalizability and
Limitations} of the methods. We \hyperref[Conclusions]{Conclude} the
paper in Chapter 5 with the main findings and policy implications.

@@

\subsection{Case Study}\label{case-study}

The Lisbon metropolitan area (LMA), in Portugal, is the case study,
comprising 18 municipalities, with a total area of 3,015 km² and a
population of 2.8 million residents, according to the 2021 census
\citep{INEcensus}. The LMA is the most populous and economically active
region in Portugal, with the city of Lisbon as the central node of the
LMA, with a high concentration of jobs and services, and a high demand
for mobility (see Figure \ref{fig:lma}).

As shown, the number of trips is proportional to the population of each
municipality, with the highest number of trips originating in Lisbon,
followed by the municipalities of Sintra, Cascais, Loures, and Almada.
The Public Transport network is well developed, although very much
Lisbon-centric, with a high level of service, including inter-municipal
trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a
metro system that serves Lisbon and surrounding municipalities.

\begin{figure}
\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to the 2021 census and the 2017 travel survey.}\label{fig:lma}
\end{figure}

The LMA is a low cycling maturity area, characterised by a sparse and
fragmented cycling network totalling 378 km, primarily concentrated
within Lisbon, and a low cycling modal share. According to the latest
mobility survey conducted in 2018 \citep{IMOB}, the LMA registered a
total of 5.3 million daily trips, with only 0.5\% by bicycle.

The survey also showed that 58.4\% of trips were made by car, 15.5\% by
public transport, and 23.9\% by walking (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
km. Nevertheless, 29\% of trips are longer than 10 km, which requires
the use of motorized modes, or active modes in combination with public
transport.

\begin{figure}
\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
\end{figure}

Regarding cycling trips in the LMA, 55\% are up to 5 km, and 88\% are up
to 10 km. These values are not in line with the typically found in
cycling distance decay curves, which show a high percentage of trips are
up to 5 km (\textasciitilde75\%), and a smaller proportion up to 10 km
\citep{krizek2007detailed, larsen2010beyond, DDfunction2023}. This may
be due to the low cycling modal share in the LMA - even more limited in
the sample survey, which may be biased towards shorter trips.

@@

Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we disaggregated the OD data into desire lines reprenting a
maximum of 100 trips each. Figure \ref{fig:jitter} illustrates the
contrast between trip representation through the traditional method,
which connects a single desire line between each district, and the
presentation achieved through the randomization and disaggregation of
trips between districts, specifically for the Lisbon metropolitan area.
As shown, the city of Lisbon (in the centre) is the main attractor of
trips, with a high number of trips to and from the other municipalities.

@@

{\centering \includegraphics[width=1\linewidth,]{img/jitter_cairo}

@@

\caption{Flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts of the 18 municipalities, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination). Furthermore, we have imposed restrictions on PT usage,
limiting it to trips without PT transfers, and within a duration of up
to 2 hours (120 minutes). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}). This map shows that some of the municipalities are not
served by PT modes that can easily accommodate bicycles, which may limit
the potential for intermodality in these areas.

This conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips \citep{vanmil_insights_2021}. Additionally,
\citet{Leferink2017} found that the bicycle is most attractive option as
a first and last mile mode when the distance ranges from 3 to 5 km, as
observed across various comparative studies. This restrictions are based
on the assumption that 5 km is a plausible maximum distance that people
are willing to travel by bike as part of a longer journey. This applies
especially for cycling beginners shifting from cars, who are less likely
to be confident and experienced cyclists than the existing cyclists.

These restrictions can be eased in the future when testing more
developed policy interventions to enhance intermodality between cycling
and PT, considering both the vehicle and infrastructure perspectives.

@@

{\centering \includegraphics[width=0.6\linewidth,]{PaperCEUS_rev1_files/figure-latex/map1-1}

@@

Figure \ref{fig:map2} illustrates the routes with greatest cycling
potential to access nodes on the public transport network. The
interactive version of the map is publicly available in the
\emph{biclaR} tool, allowing detailed exploration of local areas. At
\emph{biclaR}, users and local authority practitioners have the option
to download the results as a spatial dataset for further analysis
concerning the prioritization of cycling interventions. It is noticeable
the high potential of the train interfaces to attract car-to-PT
substituting trips.

@@

\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool). Larger the line width, higher the segment's car to bicycle shift potential. Darker the line color, lower the quietness level.}\label{fig:map2}
\end{figure}

\subsection{Modelling car shift to
bicycle-PT}\label{modelling-car-shift-to-bicycle-pt}

It is essential to note that our approach is not predictive but
scenario-based, providing a range of potential outcomes regarding the
target (\% of cycling tips) and the routing profile (safe or direct
routes), rather than a single forecast. These scenarios are rooted in
the explicit national cycling targets, assuming that the potential for
cycling is proportional to the reduction of car trips. Thus said, we do
not predict the generation of new trips, but the substitution of car
trips by cycling in combination with PT, considering the potential of
the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under
the implicit assumption that the initial and final stages of these trips
remain unaltered.

A strength of this \emph{backcasting} approach is that we can simulate
the aggregate travel patterns that would meet particular policy
objectives, and explore what the implications could be for the public
transport system and provision of safe cycling infrastructure to public
transport nodes at regional, local authority and, importantly from a
local transport planning practitioner perspective, corridor level.

@@

Following the clearly established national strategy, we estimated the
socio-environmental potential impacts of shifting car trips by bicycle
in combination with PT, in the LMA, for the two targets (4\% and 10\%).

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
daily trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles, at an aggregate level of the LMA. For the
cycling legs of the journey (first and last legs), the environmental
avoided emissions and monetized socio-environment (SE) benefits are
presented in Table \ref{tab:summary1b}, resulting from replacing car
trips with cycling.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{7em}>{\raggedleft\arraybackslash}p{7em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385\\
4\% & direct & 500 880 & 2 274 & 18 944\\
10\% & safe & 538 514 & 2 312 & 52 323\\
10\% & direct & 500 880 & 2 274 & 48 609\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

Another relevant aspect is that the higher potential (Figure
\ref{fig:mapzones}.a) is in the municipalities with the highest number
of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and
Almada, which are also the ones with the highest number of PT interfaces
(Figure \ref{fig:map1}). On another hand, the municipalities with little
PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a
lower potential for car to Bike-PT substitution (Figure
\ref{fig:mapzones}.a), even if they have a high number of trips, such as
Setubal. This findings are aligned with their local travel patterns,
where the majority of trips are intra-municipal (67\% for both
municipalities), and not to the city of Lisbon, despite the high
car-dependency (79\% in Mafra, 67\% in Setubal).

The potential and the resulting benefits are spatially unevenly
distributed across the LMA, as shown in Figure \ref{fig:mapzones}.b,
concentrating 40\% in Lisbon, where the highest number of trips are
made.

\begin{figure}
\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality (origins). The percentages are shown for municipalities with values above 4\%.}\label{fig:mapzones}
\end{figure}

Table \ref{tab:summary22} presents emissions reductions and associated
economic benefits associated with the second (PT) leg of trips. The
shift from private car associated with the PT segments would reduce
CO\textsubscript{2} equivalent emissions by 8,500 to 20,800 tons
annually, valued in €1.4 million to €3.5 million yearly, for the 4\% and
10\% targets, respectively.

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions,
highlighting the municipalities of Lisbon, Sintra, Almada, Amadora, and
Oeiras as the ones with the highest potential for car to bike-PT
substitution. Although this paper does not address the designing
specific solutions, our findings indicate a potential surge in demand
(+6.3\%) of public transport interfaces to align with the national
strategies. This added demand is conceptually driven from individuals
utilising bicycles, either personal or shared, which would be parked at
the PT stations. Consequently, operators and public space managers would
face the challenge of accommodating this increased demand for bicycle
parking and/or docks from a bike sharing system, since this availability
plays a crucial role in the decision-making process of individuals to
use combined bike-train \citep{jonkeren2021bicycle}. Additionally, they
may need to explore options to enhance bicycle carriage capacity on
public transport services, ensuring alignment with national targets.

\section{Generalizability and
Limitations}\label{generalizability-and-limitations}

This paper focus primarily on the potential for replacing trips
currently made by car with combined cycling and public transport trips,
and the associated social and environmental benefits, showcasing the
potential impact of facilitating the first-and-last mile to PT
interfaces. While the results are specific to the Lisbon Metropolitan
Area, the approach is designed to be generalizable to other metropolitan
areas.

An important aspect of the approach from a generalizability perspective
is that it has specific data requirements, including origin-destination,
road network, and the GTFS data. Such datasets are available to
transport planners in many areas. In areas where such datasets do not
exist, the value of research such as this could provide a motivation to
collect and open-up transport data. For this study we have used
Portugal's national cycling targets as the basis of the scenarios; the
approach can be adapted to other targets, or to other policy goals, such
as reducing congestion or improving air quality.

Nevertheless, the extent of the potential shift from car to PT is highly
dependent on the distances typically travelled in such contexts, and the
availability of PT services - which can vary significantly between
regions, and for which we are not assuming any change in this study. For
instance, we considered only the inter-municipal bus services that
accommodate bicycles on board. Future research could explore the
potential of equipping other regional bus services with this feature and
analyse their attractiveness, combined with cycling, to address current
car trip demands.

A strength of the approach is that there is no requirement for sensitive
individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations. The approach
cannot show exactly which trips are being replaced, in terms of the age
of people making the trips, the types of car they drive, and other
individual or vehicle level variables. We assumed uniformity in the type
of vehicle for all replaced car journeys, albeit with variations in the
fuels utilised. Additionally, we assumed consistency in the health
impacts across all individuals for the replaced car trips. This approach
is a simplification of reality to enable the analysis of a large region
on available computing resources. Future work could seek to increase the
resolution of the results, and move towards a more individual-level
analysis, for example by using agent and activity-based models using
tools such as MATSim \citep{MATSim}. This option would require
substantial resources outside the scope of this study; another way to
improve the spatial resolution of results would be to use higher
resolution `subpoints' as the basis of the desire line `jittering'
process. We would also like to explore the potential for probabilistic
routing to improve the spatial resolution of the results.

A further refinement could be to differentiate between `simpler'
no-transfer PT trips (which require less person effort) and more arduous
multi-transfer PT trips, using the same methods, by relaxing the
assumption of no-transfers between PT modes in the routing engine. This
would lead to a larger number of total daily trips that can be made by
cycling + PT, in particular in locations where the PT network is less
complex and requires more transfers. We could also set a different
maximum length of cycling stage, extending the bicycle catchment area.
However, despite the added value of a sensitivity analyses, they would
also increase the overall complexity, and would have an impact on the
explainability of the results to be used by policy makers.

\section{Conclusion}\label{Conclusions}

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA, while achieving the national cycling
targets and supporting decarbonization goals. The case study of the
Lisbon metropolitan area demonstrates that cycling-PT integration can
help meet the national targets set for bicycle use of 4\% and 10\% by
2025 and 2030, respectively.

This research also quantifies the socio-environmental benefits of
achieving such targets, exploring an intermodality scenario, where car
trips are potentially substituted by bicycle in combination with PT.

The findings indicate that cycling combined with public transport has
the potential to account for 10\% of total trips. Transitioning to
cycling for the first and last segments could lead to a decrease in
annual CO\textsubscript{2}eq emissions by a substantial margin, ranging
from 3,000 to 7,500 tons per day, depending on the cycling target and
routing profile. Moreover, for the second trip segment, the shift from
car to public transport contributes to avoiding up to 20,500 tons of
CO\textsubscript{2}eq emissions annually. These changes are estimated to
yield socio-environmental benefits totaling €125 million to €325 million
over a decade. The quantification of such benefits can support
policy-makers in prioritizing interventions to reduce the reliance on
private motorized modes of transportation.

Additionally, the results suggest that the potential for cycling-PT
intermodality is spatially unevenly distributed across the LMA. The
highest potential occurs in the municipalities with the highest number
of trips and PT interfaces with connecting modes to Lisbon that allow
for carrying bicycles on board, such as Lisbon, Sintra, Amadora, and
Almada.

We opted to detail all the steps of the methodology, from data
requirements to the final results, to ensure transparency and
generalization of the approach, for instance making sensitivity analysis
to the bike trip distances and the PT number of transfers. We
acknowledge the limitations of the used methods and propose avenues for
future research.

The presentation of the results in an open access web application will
help to inform and explain decisions. Furthermore, the provision of
datasets resulting from this project provides a foundation for further
research and development of new tools and methods. The methods are
reproducible and based on open source software, which can be applied to
other cities and metropolitan areas, supporting the decarbonization of
transport systems internationally.

>>>>

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

  \title{Modeling the impacts of replacing car trips with combined
public transport and cycling: Reproducible methods, results and
actionable evidence from biclaR}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit as a
  substitute for car trips in the Lisbon metropolitan area and assesses
  its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of
  a metropolitan cycling network was developed (\emph{biclaR}). The
  social and environmental impacts were assessed using the \emph{HEAT
  for Cycling} and the \emph{HEAT as a Service} tools. The impacts of
  shifting car trips to public transport were also estimated and
  monetized. The results show that 10\% of all trips could be made by
  cycling in combination with public transport. Shifting to cycling for
  the shorter first and last mile stages can reduce annual
  CO\textsubscript{2}eq emissions from 3,000 to 7,500 tons/day, while
  for the public transport leg, the transfer from car avoids of up to
  20,500 tons of CO\textsubscript{2}eq emissions per year. The estimated
  socio-environmental benefits are of €125 million to €325 million over
  10 years. This evidence can support policymakers to prioritize
  interventions that reduce the reliance on private motor vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326}. In The Netherlands, which has the highest
mode share of cycling in the world, cycling accounts for more than a
third of all trips to and from rail stations at the `home' end of the
journey, greatly increasing the ability of the transport system
\citep{RIETVELD200071}. This approach to reducing car dependency and
associated externalities requires interventions and programs to make
bicycling more appealing \citep{lapaix_role_2021}. The resulting public
investments can have significant social and environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high. According to the latest mobility
survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
million daily trips, with only 0.5\% by bicycle. Car modal share was
58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.\\
53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
29\% of trips are longer than 10 km, which requires the use of motorized
modes, or active modes in combination with public transport.

\begin{figure}
\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_prerev_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_prerev_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
\end{figure}

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) \citep{ENMAC},
the Lisbon's Metropolitan Department of Transport commissioned
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network \citep{felix2023}.

@@

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA, while achieving the national cycling
targets. After presenting the methods used, it assesses its
socio-environmental impacts using open data and open-source tools.

@@

@@

Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we disaggregated the OD data into desire lines reprenting a
maximum of 100 trips each. Figure \ref{fig:jitter} illustrates the
contrast between trip representation through the traditional method,
which connects a single desire line between each district, and the
presentation achieved through the randomization and disaggregation of
trips between districts, specifically for the Lisbon metropolitan area.

@@

{\centering \includegraphics[width=1\linewidth,]{img/jitter}

@@

\caption{Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination) or up to 25 minutes on bike travel-time. This
conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips \citep{vanmil_insights_2021}. Furthermore, we have
imposed restrictions on PT usage, limiting it to trips without PT
transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we have only included PT modes that can easily accommodate
bicycles, such as trains, ferries, trams, and inter-municipal bus lines
equipped with bike racks (Figure \ref{fig:map1}). These restrictions can
be eased in the future when testing more developed policy interventions
to enhance intermodality between cycling and PT, considering both the
vehicle and infrastructure perspectives.

@@

{\centering \includegraphics[width=0.6\linewidth,]{img/map1}

@@

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
the main PT interfaces in the LMA.

@@

\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
\end{figure}

@@

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are presented in Table
\ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{7em}>{\raggedleft\arraybackslash}p{7em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385\\
4\% & direct & 500 880 & 2 274 & 18 944\\
10\% & safe & 538 514 & 2 312 & 52 323\\
10\% & direct & 500 880 & 2 274 & 48 609\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:summary22} presents emissions reductions and associated
economic benefits associated with the second (PT) leg of trips. The
shift from private car associated with thes PT segments would reduce
CO\textsubscript{2} equivalent emissions by 8,500 to 20,800 tons
annually, valued in €1.4 million to €3.5 million yearly, for the 4\% and
10\% targets, respectively.

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.

\section{Conclusion}\label{conclusion}

This paper estimated the potential for combining cycling and PT to
substitute car trips in the LMA, while achieving the national cycling
targets and supporting decarbonization goals. It has become
progressively more common to establish strategic plans, at national,
regional or municipal level, to mitigate climate change. Among these,
the Sustainable Urban Mobility Plans (SUMP\footnote{See
  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
promoted by the European Commission, are becoming popular in Europe,
although authorities are designing documents of this sort all over the
world. The definition of targets associated with a timeframe for
reducing dependence on the individual motorized vehicle, or targets for
the use of active modes such as walking and cycling, are too often not
accompanied by estimates of their social, environmental and economic
impacts. It is important for authorities and practitioners to know how
to estimate those impacts, which tools are available to support them in
the process, and what results to expect.

This paper quantifies the benefits of replacing car trips with cycling,
in combination with public transport. The case study of the Lisbon
metropolitan area demonstrates that cycling-PT integration can help meet
the national targets set for bicycle use of 4\% and 10\% by 2025 and
2030, respectively.

The quantification of benefits can support policy-makers in prioritizing
interventions to reduce the reliance on private motorized modes of
transportation. The presentation of the results in an open access web
application will help to inform and explain decisions. Furthermore, the
provision of datasets resulting from this project provides a foundation
for further research and development of new tools and methods. The
methods are reproducible and based on open source software, which can be
applied to other cities and metropolitan areas, supporting the
decarbonization of transport systems internationally.

### 2024-05-09 paper/PaperCEUS/PaperCEUS_rev1.tex

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA. After presenting the
\hyperref[methods]{Methods} used in Chapter 2, we assess its
socio-environmental impacts using open data and open-source tools,
presenting the \hyperref[results-and-discussion]{Results and Discussion}
in Chapter 3, followed by Chapter 4 where we discuss the
\hyperref[generalizability-and-limitations]{Generalizability and
Limitations} of the methods. We \hyperref[Conclusions]{Conclude} the
paper in Chapter 5 with the main findings and policy implications.

>>>>

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA. After presenting the
\hyperref[methods]{Methods} used in Section 2, we assess its
socio-environmental impacts using open data and open-source tools,
presenting the \hyperref[results-and-discussion]{Results and Discussion}
in Section 3, followed by Section 4 where we discuss the
\hyperref[generalizability-and-limitations]{Generalizability and
Limitations} of the methods. We \hyperref[Conclusions]{Conclude} the
paper in Section 5 with the main findings and policy implications.
